---
output:
  html_document:
    keep_md: yes
---
#Reproducible Research: Peer Assessment 1
```{r setoptions, FALSE}
require(knitr)
opts_chunk$set( echo = FALSE , results="hide")
```
##About data
1. **Dataset**: Activity monitoring data [52K]
The variables included in this dataset are:

1. **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)

2. **date**: The date on which the measurement was taken in YYYY-MM-DD format

3. **interval**: Identifier for the 5-minute interval in which measurement was taken

##Loading and preprocessing the data
####1. Attaching required Libraries
```{r echo=TRUE}
library(stringr)
library(dplyr)
```
####2. Loading data into R
```{r echo=TRUE, results="asis"}
zip_file <- "C:/Users/Administrator/datasciencecoursera/RepData_PeerAssessment1/activity.zip"
activity_data <- read.csv(unz(zip_file,"activity.csv"),sep=",", header=T)

str(activity_data)
summary(activity_data)
head(activity_data)
```

####3. Preprocessing the data
  a. **date** field is maintaing as a **Factor**, it needs to be maintained with **Date** data type
```{r echo=TRUE}
activity_data$date <- as.Date(activity_data$date)
```
  
  b. every date observation needs to be identified as **Weekday** or **Weekend**.  Hint: Weekend day names start with **"S"**
```{r summary1, echo=TRUE, results="asis"}
activity_data$day <- weekdays(activity_data$date,abbreviate=T)
activity_data$day[substr(activity_data$day,1,1)!="S"] <- "Weekday"
activity_data$day[substr(activity_data$day,1,1)=="S"] <- "Weekend"
activity_data$day <- factor(activity_data$day)
str(activity_data)
summary(activity_data)
head(activity_data)
```

  c. 2304 observations contain *NA* entries. Further analyis does not require them. So, they need to be removed.
```{r echo=TRUE}
activity <- filter(activity_data, !is.na(steps))
```

##What is mean total number of steps taken per day?
####1. Histogram of the total number of steps taken each day
```{r plot1, echo=TRUE, fig.height=5}
steps_per_day <- summarise(group_by(activity, date), total=sum(steps))

lab_main <- "Total number of steps\ntaken per day"
lab_x    <- "Total Steps/Day"
lab_y    <- "Count of Days having those steps"

hist(steps_per_day$total,xlab=lab_x, ylab=lab_y, main=lab_main,ylim=range(1:30),labels=T, col=rainbow(5)) 
```

####2. Mean and Median total number of steps taken per day
```{r echo=TRUE, results="asis"}
v_mean   <- mean(steps_per_day$total)
v_median <- median(steps_per_day$total)
c(v_mean, v_median)
```

##What is the average daily activity pattern?
####1. Time series plot of the 5-minute interval and the average number of steps taken
```{r plot2, echo=TRUE, fig.height=5}
steps_per_interval <- summarise(group_by(activity, interval), mean=mean(steps))

lab_main <- "Average daily activity pattern"
lab_x    <- "5-minute interval (HHMM) of one Day"
lab_y    <- "Average number of Steps taken"

plot(mean~interval,data=steps_per_interval,type="l", xlab=lab_x, ylab=lab_y, main=lab_main)

y_max_val     = max(steps_per_interval$mean)
x_max_val_pos = steps_per_interval[which.max(steps_per_interval$mean),]$interval

peak_time <- paste(str_pad(x_max_val_pos%/%100,2,pad="0"),str_pad(x_max_val_pos%%100,2,pad="0"),sep=":")
peak_text = paste("Max. average steps", round(y_max_val,0), "\nare taken at", peak_time)
text(x_max_val_pos, max(steps_per_interval$mean),peak_text,pos=1, col=2)
```

####2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE, results="asis"}
print(peak_text)
```

##Imputing missing values
Replacing the missing values ( NA entries ) of original data set  **activity_data** with mean values of same interval from the data set **activity**
```{r imputed, echo=TRUE}
activity_data_imputed <- activity_data
who <- which(is.na(activity_data_imputed$steps))
f=function(x){
 g=globalenv();
 int=g$activity_data_imputed$interval[x]
 ind_df2 = which(g$steps_per_interval$interval==int)
 val=g$steps_per_interval$mean[ind_df2]
 g$activity_data_imputed$steps[x] = val
}

sapply(who, f)
```

####There are no missing values. All are replaced.
```{r echo=TRUE, results="asis"}
who <- which(is.na(activity_data_imputed$steps))
print(who)
```

##Are there differences in activity patterns between weekdays and weekends?

```{r plot4, echo=TRUE, fig.height=5}
par(mfrow=c(2,1))

activity_weekday <- filter(activity_data_imputed, day=="Weekday")
steps_per_wkday <- summarise(group_by(activity_weekday, interval), mean=mean(steps))


lab_main <- "Average Weekday activity pattern"
lab_x    <- "5-minute interval (HHMM) of one Day"
lab_y    <- "Average number of Steps taken"

plot(mean~interval,data=steps_per_wkday,type="l", xlab="",ylab=lab_y, main=lab_main)


activity_weekend <- filter(activity_data_imputed, day=="Weekend")
steps_per_wkend <- summarise(group_by(activity_weekend, interval), mean=mean(steps))

lab_main <- "Average Weekend activity pattern"
lab_x    <- "5-minute interval (HHMM) of one Day"
lab_y    <- "Average number of Steps taken"

plot(mean~interval,data=steps_per_wkend,type="l", xlab=lab_x, ylab=lab_y)
```

####Conclusion:
1. in Weekdays, more activity is during morning time 

2. in Weekends, activity is almost uniform throughout the day
